Emergent dual scaling of riverine biodiversity

### 2021-06-30 manuscript_pnas.Rmd
edits: main text
Update manuscript_pnas.Rmd

---
title: "Emergent dual scaling of riverine biodiversity"
output:
  word_document:
    reference_docx: word_format.docx
csl: pnas.csl
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile, encoding = encoding, output_dir = "document_output")
      })
bibliography: references.bib
---

@@

**Authors:** Akira Terui^1^, Seoghyun Kim^1^, Christine L. Dolph^2^, Taku Kadoya^3^, Yusuke Miyazaki^4^

@@

$r_{ix}(t)$ is the site- and time-specific reproductive rate of species $i$, $r_{0,i}$ the maximum reproductive rate, $K_{x}$ the site-specific carrying capacity, $b_{ij}$ the competition coefficient between species $i$ and $j$, and $S$ the number of species in a metacommunity ($S = 50$). Interspecific competition coefficients are random draws from a uniform distribution as $b_{ij, i \neq j} \sim Unif(0,b_{max})$ while keeping intraspecific competition coefficients constant ($b_{ii} = 1$). Here, the reproductive rate $r_{ix}(t)$ was related to local abiotic environments through a Gaussian function with species' niche optimum $\mu_i$ and width $\sigma_{niche,i}$. That is, $r_{ix}(t)$ decreases as temporally-variable abiotic environments (drawn randomly from the mean environment $\mu_{z,x}$) deviate from $\mu_i$ (see **Materials and Methods**). Species' reproductive performance has therefore an explicit linkage with local abiotic environments, which can be driven largely by branching structure. After local community dynamics, species emigrate from the local habitat patch with probability $p_d$ and disperse through the branching network following an exponential dispersal kernel with the rate parameter $\theta$ (the reciprocal $1/\theta$ represents an expected dispersal distance). These dispersal parameters affect the number of immigrants $I_{ix}(t)$ and emigrants $E_{ix}(t)$.

@@

The consistent effect of branching probability on $\gamma$ diversity across the study regions is noteworthy because many watersheds in the Midwest have been altered by agricultural land use (mean % agricultural land use: `r round(mean(wsd_subset[[2]]$frac_agri)*100)`% in the Midwest and `r round(mean(wsd_subset[[1]]$frac_agri)*100)`% in Hokkaido). If the intensive land use by humans impairs biological or physical distinctiveness among tributaries, theory predicts a weakened effect of branching probability on $\gamma$ diversity (**Figures S13-S18**). However, $\gamma$ diversity increased significantly with increasing branching probability in this highly modified landscape, suggesting that tributaries still sustain unique environmental conditions to support high spatial variation in species composition. Indeed, $\beta$ diversity increased with increasing branching probability in both regions (**Figure 4**). It is conceivable that local geological and geomorphological differences, such as slope, aspect, and soil porosity, persist in human-dominated landscapes to maintain the diversity of in-stream processes. The lack of land use effects (the fraction of agricultural land use and dam density) further corroborates our interpretation (**Table 1**). Although our analysis is correlative, the finding is encouraging because the branching complexity of river networks may serve as a natural defense system to human-induced environmental changes.

While our empirical results showed an excellent match with the predictions of our niche-based model (**Figures 2 and 3**), we must note that a neutral model developed by Muneepeerakul et al. [@muneepeerakulNeutralMetacommunityModels2008a] was also capable of reproducing fish diversity patterns in the Mississippi-Missouri River Basin. This apparent contradiction is perhaps not a surprise when considering the difference in the spatial resolution of interest. Their neutral model used sub-basins (ca. 5000 km^2^ in the watershed area) as a unit of local communities (metacommunities in our study) and modeled variation in habitat capacity and dispersal among those units. At this coarse resolution, stochastic immigration processes coupled with random speciation, a core of the Unified Neutral Theory of Biodiversity [@hubbellUnifiedNeutralTheory2001], are critical to forming biogeographic patterns [@carvajal-quinteroDrainageNetworkPosition2019]. Meanwhile, we defined local communities at much finer resolutions (several tens to hundred meters) at which species' niche should interact with environments to determine community structure [@jacksonWhatControlsWho2001]. It is therefore necessary for our theoretical model to assume niche-based processes. This interpretation is supported by the fact that a neutral model could not reproduce the positive complexity-diversity relationship observed in our empirical data [@muneepeerakulEvolvingBiodiversityPatterns2019]. As such, our research is uniquely significant in showing the qualitative agreement of theoretical and empirical results at this level of biological organization.

@@

While prevailing evidence supports the importance of ecosystem size in biodiversity relationships [@hubbellUnifiedNeutralTheory2001; @triantisIslandSpeciesareaRelationship2012; @iwasakiEvaluatingRelationshipBasinscale2012; @saboRoleDischargeVariation2010], ecosystem complexity has not received the attention it deserves. However, the ubiquity of scale-invariant complexity across terrestrial [@turnerLandscapeEcologyTheory2015] and aquatic ecosystems [@rodriguez-iturbeFractalRiverBasins2001] calls for more research embracing the two orthogonal dimensions of the ecosystem's geometric structure. Our fundamental findings should apply beyond branching ecosystems because ecosystem size and complexity both represent a physical 'template' providing a wide spectrum of niche opportunities for living organisms. Further, the proposed framework may be extended to other aspects of biodiversity; for example, a logical next step is to explore how these ecosystem properties regulate functional diversity [@petcheyFunctionalDiversityFD2002; @suHumanImpactsGlobal2021] and shifts in community assembly processes [@chaseStochasticCommunityAssembly2010]. Recognizing the dual scaling of biodiversity broadens viable options for spatial planning of protected areas or restoration sites, thereby helping conserve biodiversity from societal demands that threaten it. The generalization of our findings in spatially complex ecosystems represents a frontier for future research.

@@

We considered watersheds as independent replicates of metacommunities if the river network flows into one of the following: the ocean, a large lake/reservoir ($\ge$ 10 km² in the areal area) or a large river that may represent lentic habitats ($\ge$ 5000 km² in the watershed area). We estimated $\alpha$, $\beta$, and $\gamma$ diversity at each watershed. $\gamma$ diversity is the total species richness at each watershed. Since it is impossible to sample all species present in each watershed, we used asymptotic species richness as a proxy for $\gamma$ diversity. Asymptotic species richness can be interpreted as the estimated true species richness in a given area when estimated using spatial replicates of local community data within a metacommunity [@chaoRarefactionExtrapolationHill2014a]. We estimated asymptotic species richness using sampling sites located in the same watershed (i.e., spatial replicates within a metacommunity) via the R package 'iNEXT' [@hsiehINEXTPackageRarefaction2016]. Metacommunities with $\ge$ 10 sampling sites were used in our analysis because a small sample size inflates statistical uncertainties in estimating asymptotic species richness (**Supporting Information**). We assessed the completeness of the observed species richness based on sample coverage, which is the total proportional abundances of observed species to the whole metacommunity [@chaoRarefactionExtrapolationHill2014a]. We excluded watersheds whose sample coverage was below 0.9 because the estimates of asymptotic species richness may contain non-negligible statistical uncertainty. $\alpha$ diversity was estimated by taking an average of local species richness at each site. $\beta$ diversity was estimated by dividing the estimated $\gamma$ diversity by $\alpha$ diversity [@whittakerVegetationSiskiyouMountains1960].

@@

We are grateful to numerous people who have been involved in the collection of fish community data. We thank Michio Fukushima and Hirokazu Urabe for providing information on the Hokkaido Freshwater Fish Database HFish and the fish monitoring data at protected watersheds in Hokkaido. We thank Jacques Finlay for his constructive comments on the earlier version of the manuscript. This material is based upon work supported by the National Science Foundation through the Water Sustainability and Climate Program Observatory grant (EAR‐1209402): REACH (Resilience under Accelerated Change) and through the Division of Environmental Biology (DEB 2015634), Environmental Protection Agency through the Water Quality Benefits grant (EPA-G2015-STAR-A1), the University of North Carolina at Greensboro through a start-up fund, and the Japan Society for the Promotion of Science through the KAKENHI grant (20H03010).

>>>>

---
title: "Emergent dual scaling of riverine biodiversity"
output:
  word_document:
    reference_docx: word_format.docx
csl: pnas.csl
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile, encoding = encoding, output_dir = "document_output")
      })
bibliography: reference.bib
---

@@

**Authors:** Akira Terui^1,\*^, Seoghyun Kim^1^, Christine L. Dolph^2^, Taku Kadoya^3^, Yusuke Miyazaki^4^

\*Corresponding author: hanabi0111\@gmail.com

@@

$r_{ix}(t)$ is the site- and time-specific reproductive rate of species $i$, $r_{0,i}$ the maximum reproductive rate, $K_{x}$ the site-specific carrying capacity, $b_{ij}$ the competition coefficient between species $i$ and $j$, and $S$ the number of species in a metacommunity ($S = 50$). Interspecific competition coefficients are random draws from a uniform distribution as $b_{ij, i \neq j} \sim Unif(0,b_{max})$ while keeping intraspecific competition coefficients constant ($b_{ii} = 1$). Here, the reproductive rate $r_{ix}(t)$ was related to local abiotic environments through a Gaussian function with species' niche optimum $\mu_i$ and width $\sigma_{niche,i}$. That is, $r_{ix}(t)$ decreases as the temporally-variable abiotic environment $z_x(t)$ deviates from $\mu_i$ ($z_x(t)$ is drawn randomly from a multivariate normal distribution with the mean environment of the patch $\mu_{z,x}$; see **Materials and Methods**). Species' reproductive performance has therefore an explicit linkage with local abiotic environments, which can be driven largely by branching structure. After local community dynamics, species emigrate from the local habitat patch with probability $p_d$ and disperse through the branching network following an exponential dispersal kernel with the rate parameter $\theta$ (the reciprocal $1/\theta$ represents an expected dispersal distance). These dispersal parameters affect the number of immigrants $I_{ix}(t)$ and emigrants $E_{ix}(t)$.

@@

The consistent effect of branching probability on $\gamma$ diversity across the study regions is noteworthy because many watersheds in the Midwest have been altered by agricultural land use (mean % agricultural land use: `r round(mean(wsd_subset[[2]]$frac_agri)*100)`% in the Midwest and `r round(mean(wsd_subset[[1]]$frac_agri)*100)`% in Hokkaido). If the intensive land use by humans impairs biological or physical distinctiveness among tributaries, our theory predicts a weakened effect of branching probability on $\gamma$ diversity (**Figures S13-S18**). However, $\gamma$ diversity increased significantly with increasing branching probability in this highly modified landscape, suggesting that tributaries still sustain unique environmental conditions to support high spatial variation in species composition. Indeed, $\beta$ diversity increased with increasing branching probability in both regions (**Figure 4**). It is conceivable that local geological and geomorphological differences, such as slope, aspect, and soil porosity, persist in human-dominated landscapes to maintain the diversity of in-stream processes. The lack of land use effects (the fraction of agricultural land use and dam density) further corroborates our interpretation (**Table 1**). Although our analysis is correlative, the finding is encouraging because the branching complexity of river networks may serve as a natural defense system to human-induced environmental changes.

While our empirical results showed an excellent match with the predictions of our niche-based model (**Figures 2 and 3**), we must note that a neutral model developed by Muneepeerakul et al. [@muneepeerakulNeutralMetacommunityModels2008] was also capable of reproducing fish diversity patterns in the Mississippi-Missouri River Basin. This apparent contradiction is perhaps not a surprise when considering the difference in the spatial resolution of interest. Their neutral model used sub-basins (ca. 5000 km^2^ in the watershed area) as a unit of local communities (metacommunities in our study) and modeled variation in habitat capacity and dispersal among those units. At this coarse resolution, stochastic immigration processes coupled with random speciation, a core of the Unified Neutral Theory of Biodiversity [@hubbellUnifiedNeutralTheory2001], are critical to forming biogeographic patterns [@carvajal-quinteroDrainageNetworkPosition2019]. Meanwhile, we defined local communities at much finer resolutions (several tens to hundred meters) at which species' niche should interact with environments to determine community structure [@jacksonWhatControlsWho2001]. It is therefore necessary for our theoretical model to assume niche-based processes. This interpretation is supported by the fact that a neutral model could not reproduce the positive complexity-diversity relationship observed in our empirical data [@muneepeerakulEvolvingBiodiversityPatterns2019]. As such, our research is uniquely significant in showing the qualitative agreement of theoretical and empirical results at this level of biological organization.

@@

While prevailing evidence supports the importance of ecosystem size in biodiversity relationships [@hubbellUnifiedNeutralTheory2001; @triantisIslandSpeciesareaRelationship2012; @iwasakiEvaluatingRelationshipBasinscale2012; @saboRoleDischargeVariation2010], ecosystem complexity has not received the attention it deserves. However, the ubiquity of scale-invariant complexity across terrestrial [@turnerLandscapeEcologyTheory2015] and aquatic ecosystems [@rodriguez-iturbeFractalRiverBasins2001] calls for more research embracing the two orthogonal dimensions of the ecosystem's geometric structure. Our fundamental findings should apply beyond branching ecosystems because ecosystem size and complexity both represent a physical 'template' providing a wide spectrum of niche opportunities for living organisms. Further, the proposed framework may be extended to other aspects of biodiversity; for example, a logical next step is to explore how these ecosystem properties regulate functional diversity [@petcheyFunctionalDiversityFD2002; @suHumanImpactsGlobal2021] and shifts in stochastic/deterministic community assembly [@chaseStochasticCommunityAssembly2010]. Recognizing the dual scaling of biodiversity broadens viable options for spatial planning of protected areas or restoration sites, thereby helping conserve biodiversity from societal demands that threaten it. The generalization of our findings in spatially complex ecosystems represents a frontier for future research.

@@

We considered watersheds as independent replicates of metacommunities if the river network flows into one of the following: the ocean, a large lake/reservoir ($\ge$ 10 km² in the areal area) or a large river that may represent lentic habitats ($\ge$ 5000 km² in the watershed area). We estimated $\alpha$, $\beta$, and $\gamma$ diversity at each watershed. $\gamma$ diversity is the total species richness at each watershed. Since it is impossible to sample all species present in each watershed, we used asymptotic species richness as a proxy for $\gamma$ diversity. Asymptotic species richness can be interpreted as the estimated true species richness in a given area when estimated using spatial replicates of local community data within a metacommunity [@chaoRarefactionExtrapolationHill2014]. We estimated asymptotic species richness using sampling sites located in the same watershed (i.e., spatial replicates within a metacommunity) via the R package 'iNEXT' [@hsiehINEXTPackageRarefaction2016]. Metacommunities with $\ge$ 10 sampling sites were used in our analysis because a small sample size inflates statistical uncertainties in estimating asymptotic species richness (**Supporting Information**). We assessed the completeness of the observed species richness based on sample coverage, which is the total proportional abundances of observed species to the whole metacommunity [@chaoRarefactionExtrapolationHill2014]. We excluded watersheds whose sample coverage was below 0.9 because the estimates of asymptotic species richness may contain non-negligible statistical uncertainty. $\alpha$ diversity was estimated by taking an average of local species richness at each site. $\beta$ diversity was estimated by dividing the estimated $\gamma$ diversity by $\alpha$ diversity [@whittakerVegetationSiskiyouMountains1960].

@@

We are grateful to numerous people who have been involved in the collection of fish community data. We thank Michio Fukushima and Hirokazu Urabe for providing information on the Hokkaido Freshwater Fish Database HFish and the fish monitoring data at protected watersheds in Hokkaido. We thank Jacques Finlay for his constructive comments on the earlier version of the manuscript. This material is based upon work supported by the National Science Foundation through the Water Sustainability and Climate Program Observatory grant (EAR‐1209402): REACH (Resilience under Accelerated Change) and through the Division of Environmental Biology (DEB 2015634), Environmental Protection Agency through the Water Quality Benefits grant (EPA-G2015-STAR-A1), the University of North Carolina at Greensboro through start-up funds, and the Japan Society for the Promotion of Science through the KAKENHI grant (20H03010).
